Assessment and Optimization of Nowcast Measures for Trend Detection

### 2024-01-13 paper.tex
Update on Overleaf.

@@

%%% --- Colors
\usepackage[]{xcolor}

@@

\begin{itemize}
    \item Welche Form sollte das Maß haben? Mögliche Wahlen: Indikator, Probabilistisch, Weighted (1), Weighted (2), Indikator mit Korrektur, ... (Siehe Tabelle \ref{tab:comparison-measures})
    \item Schätzung/Wahl des Rauschparameters $\sigma$; haben wir da irgendeine Methode aus den Daten das Sigma zu bekommen? Filter vs. Empirie 
    \item Gehen wir auf andere Verteilungen für den Fehler ein?
    \item Konfidenzintervalle
\end{itemize}

@@

@@

\begin{table}[]
    \centering
    \begin{tabular}{l l p{4cm} p{4cm} p{4cm}}
    \toprule
    name & abbr. & accuracy & + & - \\ \midrule
    indicator & i & $\frac{1}{T-1} \sum \ind{x_t y_t > 0}$ &  well-known, intuitive & what about points near (0,0) \\
    Probabilistic & p & $\frac{1}{T-1} \sum P{(x_t y_t > 0)}$ & small variance, weighting due to probabilities & unintuitive, asymptotics unclear, driven towards 0.5, no standard estimation for $\sigma$  \\
    weighted 1 (*) & w1 & {\small{$\sum w_i \ind{x_t y_t > 0} / \sum w_i$,\newline $w_i = \max_j\{P(y_i, x_i) \in Q_j \}$}} & intuitive, implicit exclusion area, natural extension & all points gain weight, estimation of $\sigma$. \\
    weighted 2 (*) & w2 & {\small {$\sum w_i \ind{x_t y_t > 0} / \sum w_i$,\newline $w_i = \max_j\{P(y_i, x_i) \in Q_j \} - 0.25$}} & as w1, points near (0,0) do not gain weight & \\ \bottomrule
    \end{tabular}
    \caption{Comparison between different measures for accuracy. }
    \label{tab:comparison-measures}
\end{table}

\unsure[inline]{Anpassung der gewichteten Formel}
(*): The weighted formula can also be changed by using the signum function $\sign$ instead of $\ind$ to take into account the influence of negative trending.\\

Natural evaluation method for 4Q-plot, as known from dichotomous event evaluation, evaluation of classification (confusion matrix)

\begin{itemize}
  \item Accuracy: \begin{equation}
  	m_{\text{acc}} = \frac{\sum_{t=T-w}^T \ind{\ydiff_t > 0, \xdiff_t > 0} + \sum_{t=T-w}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{T-w}
\end{equation}
	\item Capability of detecting a positive trend (probability of detection)/Sensitivity
	\begin{equation}
  		m_{\text{py}} = \frac{\sum_{t=T-w}^T \ind{\ydiff_t > 0, \xdiff_t > 0}}{\sum_{t=T-w}^T \ind{\ydiff_t > 0}}
	\end{equation}
	\item positive predictive values
\begin{equation}
  m_{\text{px}} = \frac{\sum_{t=T-w}^T \ind{\ydiff_t > 0, \xdiff_t > 0}}{\sum_{t=T-w}^T \ind{\xdiff_t > 0}}  
\end{equation}
	\item Specificity
	\begin{align}
		m_{\text{ny}} &=  \frac{\sum_{t=T-w}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{\sum_{t=T-w}^T \ind{\ydiff_t < 0}} 
  \end{align}
  \item Negative predictive value
	\begin{align}
		m_{\text{nx}} &=  \frac{\sum_{t=T-w}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{\sum_{t=T-w}^T \ind{\xdiff_t < 0}}
	\end{align}
\end{itemize}

@@

Assume an additive error decomposition for both observation and nowcast to account for non-systematic and short-term (i.e., intra-day) effects or for effects on inaccurate representation of $y_{t-1}$ in the computation of $Dx_t$ (\hl{this, no additional complexity}):
  	\begin{align}\label{additive error decomposition}
  		\ydiff_t &= \ydifft + \varepsilon_t^y \\
  		\xdiff_t &= \xdifft + \varepsilon_t^x
	\end{align}

Extension 1: 
\begin{itemize}
  \item 
Adapt the measures to be weighted by the explanatory power of the points:\\
Weights $\omega_t$ for $(\ydiff_t, \xdiff_t)$ could be determined by error model and corresponds to the maximal likeliness of the point belonging to the different quadrants ($Q_1, \dots, Q_4$); thus, point that have less explanatory power for assessing the trending ability becomes less weight:
 	\begin{equation}
  		\omega_t = \max_{i \in [4]} {P\big( (\ydifft_t, \xdifft_t) \in Q_i \cond \ydiff_t, \xdiff_t \big)} 
	\end{equation}
	\item E.g., the accuracy becomes
	\begin{equation}
  m_{acc, \omega} = \frac{\sum_{t=T-w}^T \omega_t \ind{\ydiff_t \cdot \xdiff_t > 0} }{\sum_{t=T-w}^T \omega_t}
\end{equation}
	\item Thus, $\omega_t \in [0.25, 1]$. Maybe it would make sense to \enquote{ignore} values without explanatory power and use $\omega_t^\star = 4/3 \ \omega_t \in [0, 1]$ \hl{w2}
	\item Figure~\ref{fig:comparison_estimators} shows the resulting estimates for the simulation study comparing the pure indicator, the probability-version and the weighted indicator estimation; the data is generated (in a very simpel model) by
	\begin{align}
		\ydifft_t &\sim N(0, 2) \\
		\xdifft_t &= (2 b_t - 1) \cdot \ydiff_t , b_t \sim \Ber (k) \\
		\ydiff_t &= \ydifft_t + \varepsilon_t^y, \varepsilon_t^y \sim N(0, 0.1) \\
		\xdiff_t &= \xdifft_t + \varepsilon_t^x, \varepsilon_t^x \sim N(0, 0.1) \\
	\end{align}
\end{itemize}

\begin{figure}
  \centering
	\includegraphics{plots/unbiasedness/boxplot_k_0.1.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.3.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.5.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.7.pdf}
	\includegraphics{plots/unbiasedness/boxplot_k_0.9.pdf}			
  \caption{Simulations results comparing the three types of estimators for the accuracy in a trending setting}
  \label{fig:comparison_estimators}
\end{figure}

Probabilistic extension:
\begin{itemize}
\item Probabilistic versions of non-probabilistic measures; small sample characteristic of $m_{acc}$ can be driven by random effects near the axis; therefore probabilistic \enquote{correction}:
\begin{itemize}
  \item Assume an additive error decomposition for both observation and nowcast to account for non-systematic and short-term (i.e., intra-day) effects or for effects on inaccurate representation of $y_{t-1}$ in the computation of $Dx_t$ (\hl{this, no additional complexity}):
  	\begin{align}\label{additive error decomposition}
  		\ydiff_t &= \ydifft + \varepsilon_t^y \\
  		\xdiff_t &= \xdifft + \varepsilon_t^x
	\end{align}
 \item Wouldn't it be more natural to model the original time series with a noise term? (Qualitatively, it would yield the same results)
 \begin{align}
     y_t &= y_t^\star + \epsilon_t^y \\
     x_t &= x_t^\star + \epsilon_t^x \\
     \ydiff_t &= y_t^\star + \epsilon_t^y - (y_{t-1}^\star + \epsilon_{t-1}^y) = y_t^\star - y_{t-1}^\star + \underbrace{\epsilon_t^y - \epsilon_{t-1}^y}_{\sim N(0, \sqrt{2} \sigma_y)} \\
     \xdiff_t &= x_t^\star + \epsilon_t^x - (x_{t-1}^\star + \epsilon_{t-1}^x) = x_t^\star - x_{t-1}^\star + \underbrace{\epsilon_t^x - \epsilon_{t-1}^x}_{\sim N(0, \sqrt{2} \sigma_x)}
 \end{align}
 Advantage of the approach: For different lags, no new standard deviation has to be estimated.\\
 Disadvantage: Theoretically, there is serial correlation of the errors. (Wie behandle ich die Autokorrelation; wie rausrechnen)
\item  Replace indicators in the above formulations by the corresponding probabilities:
\begin{align}
		\mprob_{\text{acc}} &= \frac{\sum_{t=T-w}^T \Prob{ \ydifft_t > 0, \xdifft_t > 0} + \sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{T-w}  \\
   \mprob_{\text{py}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t > 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t > 0}} \\
    \mprob_{\text{ny}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t < 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t < 0}} 
\end{align} 
\item Simple error model:
  \begin{align}
	  \varepsilon_Y \sim N(0, \sigma_Y) \\
	  \varepsilon_X \sim N(0, \sigma_X)
  \end{align}
  yields, e.g.,
  	\begin{equation}
  		\mprob_{\text{acc}} = \frac{1}{T-w} \sum_{t=T-w}^T  \big( 1 - \Phi_{\ydiff_t, \sigma_y}(0) - \Phi_{\xdiff_t, \sigma_x} (0) + 2 \Phi_{\xdiff_t,\sigma_x}( 0)\Phi_{\ydiff_t,\sigma_y}( 0) \big), 
	\end{equation}
	where $\Phi$ denotes a (possibly multivariate) normal distribution
\end{itemize}
\item For $\sigma_x \sigma_y \rightarrow 0$, the probabilistic measures approach the count-based ones
\item For $\sigma_x \sigma_y \rightarrow \infty$, the probabilistic accuracy approaches   0.5 as all probabilities are approximately 0.5
\item Lag-$l$-measures can easily be obtained by using the $D^l$-difference, i.e., $D^l y_t = y_t - y_{t-l}$
\end{itemize}

\unsure[inline]{Diskussion der Annahme der additiven Fehler-Dekomposition}

Nach Annahme der additiven Fehler-Dekomposition ist es ohne weiteres nicht möglich, die probabilistische Formel, insbesondere den Ausdruck $P(Dy^*\cdot Dx^* > 0|Dx=Dx_t, Dy=Dy_t) $ auszurechnen, da $(Dx,Dy)$ und $(Dx^*,Dy^*)$ abhängig sind

\subsection{Gewichtete Formel }
\unsure[inline]{Gewichtete Formel mit einer anderen Modelannahme und storyline}

@@

>>>>

%%% --- Colors
\usepackage[dvipsnames]{xcolor}

@@

@@

\begin{itemize}
    \item Welche Form sollte das Maß haben? Mögliche Wahlen: Indikator, Probabilistisch, verschiedene Gewichtungen, ... (Siehe Tabelle \ref{tab:comparison-measures})
    \item Indikator oder Signum?
    \item Schätzung/Wahl des Rauschparameters $\sigma$ bei Maßen außer Indikator
    \item Gehen wir auf andere Verteilungen für den Fehler ein?
    \item Konfidenzintervalle
\end{itemize}

@@

Comparsion and outline of various ideas for measures; overview in Table~\ref{tab:comparison-measures}:
\begin{itemize}
    \item Subsection~\ref{sec:equal-weight} Equal-weighted indicator
    \item Subsection~\ref{subsec:w1} Weighted formula by explanatory power of points
    \item Subsection~\ref{subsec:w2} Combined assessment of trend and its estimation uncertainty with weighting by explanatory power of points
    \item A comparison of the Measures of Subsection~\ref{subsec:w1} and Subsection~\ref{subsec:w2} is given in \ref{subsec:w1-w2}
    \item Subsection~\ref{subsec:probabilistic} Probabilistic version of estimation (as historic version)
\end{itemize}

@@

\begin{table}[]
    \centering
    \begin{tabular}{l l p{4cm} p{4cm} p{4cm}}
    \toprule
    name & abbr. & accuracy & + & - \\ \midrule
    indicator & i & $\frac{1}{T-1} \sum \ind{x_t y_t > 0}$ &  well-known, intuitive & what about points near (0,0) \\
    Probabilistic & p & $\frac{1}{T-1} \sum P{(x_t y_t > 0)}$ & small variance, weighting due to probabilities & unintuitive, asymptotics unclear, driven towards 0.5, no standard estimation for $\sigma$  \\
    weighted 1 (*) & w1 & {\small {$\sum w_i \ind{x_t y_t > 0} / \sum w_i$,\newline $w_i = \max_j\{P(y_i, x_i) \in Q_j \} - 0.25$}} & intuitive, natural extension, implicit exclusion area as points near (0,0) do not gain weight & \\ \bottomrule
    \end{tabular}
    \caption{Comparison between different measures for accuracy. For the weighted measures, the indicator could be replaced by the signum function of the product.}
    \label{tab:comparison-measures}
\end{table}

\subsubsection{Equal-weighted indicators} \label{sec:equal-weight}

Natural evaluation method for 4Q-plot, as known from dichotomous event evaluation, evaluation of classification (confusion matrix):

\begin{tabularx}{\textwidth}{p{5cm} X}
     Accuracy & $m_{\text{acc}} = \frac{\sum_{t=T-w}^T \ind{\ydiff_t > 0, \xdiff_t > 0} + \sum_{t=T-w}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{T-w}$ \\
     Capability of detecting a positive trend (probability of detection)/Sensitivity & $m_{\text{py}} = \frac{\sum_{t=T-w}^T \ind{\ydiff_t > 0, \xdiff_t > 0}}{\sum_{t=T-w}^T \ind{\ydiff_t > 0}}$ \\
     Positive predictive values & $ m_{\text{px}} = \frac{\sum_{t=T-w}^T \ind{\ydiff_t > 0, \xdiff_t > 0}}{\sum_{t=T-w}^T \ind{\xdiff_t > 0}}$ \\
     Specificity & $m_{\text{ny}} =  \frac{\sum_{t=T-w}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{\sum_{t=T-w}^T \ind{\ydiff_t < 0}} $ \\
     Negative predictive value & $m_{\text{nx}} =  \frac{\sum_{t=T-w}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{\sum_{t=T-w}^T \ind{\xdiff_t < 0}}$
\end{tabularx}

@@

% Extension 1: 
% \begin{itemize}
%   \item 
% Adapt the measures to be weighted by the explanatory power of the points:\\
% Weights $\omega_t$ for $(\ydiff_t, \xdiff_t)$ could be determined by error model and corresponds to the maximal likeliness of the point belonging to the different quadrants ($Q_1, \dots, Q_4$); thus, point that have less explanatory power for assessing the trending ability becomes less weight:
%  	\begin{equation}
%   		\omega_t = \max_{i \in [4]} {P\big( (\ydifft_t, \xdifft_t) \in Q_i \cond \ydiff_t, \xdiff_t \big)} 
% 	\end{equation}
% 	\item E.g., the accuracy becomes
% 	\begin{equation}
%   m_{acc, \omega} = \frac{\sum_{t=T-w}^T \omega_t \ind{\ydiff_t \cdot \xdiff_t > 0} }{\sum_{t=T-w}^T \omega_t}
% \end{equation}
% 	\item Thus, $\omega_t \in [0.25, 1]$. Maybe it would make sense to \enquote{ignore} values without explanatory power and use $\omega_t^\star = 4/3 \ \omega_t \in [0, 1]$ \hl{w2}
% 	\item Figure~\ref{fig:comparison_estimators} shows the resulting estimates for the simulation study comparing the pure indicator, the probability-version and the weighted indicator estimation; the data is generated (in a very simpel model) by
% 	\begin{align}
% 		\ydifft_t &\sim N(0, 2) \\
% 		\xdifft_t &= (2 b_t - 1) \cdot \ydiff_t , b_t \sim \Ber (k) \\
% 		\ydiff_t &= \ydifft_t + \varepsilon_t^y, \varepsilon_t^y \sim N(0, 0.1) \\
% 		\xdiff_t &= \xdifft_t + \varepsilon_t^x, \varepsilon_t^x \sim N(0, 0.1) \\
% 	\end{align}
% \end{itemize}

% \begin{figure}
%   \centering
% 	\includegraphics{plots/unbiasedness/boxplot_k_0.1.pdf}
% 	\includegraphics{plots/unbiasedness/boxplot_k_0.3.pdf}
% 	\includegraphics{plots/unbiasedness/boxplot_k_0.5.pdf}
% 	\includegraphics{plots/unbiasedness/boxplot_k_0.7.pdf}
% 	\includegraphics{plots/unbiasedness/boxplot_k_0.9.pdf}			
%   \caption{Simulations results comparing the three types of estimators for the accuracy in a trending setting}
%   \label{fig:comparison_estimators}
% \end{figure}

\subsubsection{Weighted measure by accounting for explanatory power of the points}\label{subsec:w1}

Unlike the situation in dichotomous forecast evaluation or binary classification, the points $(\ydiff_t, \xdiff_t)$ are not binary.
For points with small $\ydiff_t$ and $\xdiff_t$, the sign of their product is more likely to be driven by noise and not a systematic relationship.
Weighting the observations by their likelihood to be driven by noise makes the estimation more robust and does not decrease the trending ability because of times of small and nonsystematic changes. 
A simple method to account for the uncertainty in the measurement of $\ydiff_t$ and $\xdiff_t$ is to assume that the true values $\ydifft_t$ and $\xdifft$ follow (independent) Gaussian distributions around their observed counterparts with standard deviation $\sigma_y$ and $\sigma_y$, respectively.
Then, the estimation can be weighted by how clearly a point can be assigned to one of the quadrants:
\begin{equation}
    m_{\text{acc}}^1 = \frac{\sum_{t=T-w}^T \omega_t \ind{\ydiff_t \xdiff_t > 0} }{\sum w_t}
\end{equation}
with $w_t = \max_{i \in [4]} P((\ydiff_t \xdiff_t) \in Q_i) - 0.25$.
Thus, the $w_t$ are in $[0, 0.75]$ and points very close to $(0, 0)$ do not have any weight in assessing the trend. 
Due to the typical weight structure, $m_{\text{acc}}^1$ is in $[0,1]$ and the bounds are attained if and only if all points are concordant or disconcordant, respectively.

Instead of using the indicator of concordance, one could also use the signum function of the product of $(\ydiff_t, \xdiff_t)$ to scale the result to $(-1, 1)$.

\subsubsection{Combined measure of trending and its uncertainty} \label{subsec:w2}
\unsure[inline]{Gewichtete Formel mit einer anderen Modelannahme und storyline}

@@

\newpage
\subsubsection{Comparison of weighted measures using examples} \label{subsec:w1-w2}

\begin{tabularx}{\textwidth}{c l l l}
     & Indicator & Weighted 1 & Weighted 2 \\
     Verrauschte Daten & & & \\
     Konzentration in Q1/Q3, aber viel an Entscheidungsgrenzen, kein Noise & & & \\
     $\ydiff_t = \xdiff_t$ und $\ydiff_t \sim N(0, 1)$ & & & \\

\end{tabularx}

\newpage
\subsubsection{Probabilistic measure (wrong derivation)} \label{subsec:probabilistic}

\improvement[inline]{Due to the additive structure, $(Dx,Dy)$ and $(Dx^*,Dy^*)$ are not independet, and $P(Dy^*\cdot Dx^* > 0|Dx=Dx_t, Dy=Dy_t) $ is difficult to compute}

\begin{itemize}
\item Probabilistic versions of non-probabilistic measures; small sample characteristic of $m_{acc}$ can be driven by random effects near the axis; therefore probabilistic \enquote{correction}:
\begin{itemize}
  \item Assume an additive error decomposition for both observation and nowcast to account for non-systematic and short-term (i.e., intra-day) effects or for effects on inaccurate representation of $y_{t-1}$ in the computation of $Dx_t$ (\hl{this, no additional complexity}):
  	\begin{align}\label{additive error decomposition}
  		\ydiff_t + \varepsilon_t^y &= \ydifft  \\
  		\xdiff_t + \varepsilon_t^x &= \xdifft 
	\end{align}
\item  Replace indicators in the above formulations by the corresponding probabilities:
\begin{align}
		\mprob_{\text{acc}} &= \frac{\sum_{t=T-w}^T \Prob{ \ydifft_t > 0, \xdifft_t > 0} + \sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{T-w}  \\
   \mprob_{\text{py}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t > 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t > 0}} \\
    \mprob_{\text{ny}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\ydifft_t < 0}} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=T-w}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=T-w}^T \Prob{\xdifft_t < 0}} 
\end{align} 
\item Simple error model:
  \begin{align}
	  \varepsilon_Y \sim N(0, \sigma_Y) \\
	  \varepsilon_X \sim N(0, \sigma_X)
  \end{align}
  yields, e.g.,
  	\begin{equation}
  		\mprob_{\text{acc}} = \frac{1}{T-w} \sum_{t=T-w}^T  \big( 1 - \Phi_{\ydiff_t, \sigma_y}(0) - \Phi_{\xdiff_t, \sigma_x} (0) + 2 \Phi_{\xdiff_t,\sigma_x}( 0)\Phi_{\ydiff_t,\sigma_y}( 0) \big), 
	\end{equation}
	where $\Phi$ denotes a (possibly multivariate) normal distribution
\end{itemize}
\item For $\sigma_x \sigma_y \rightarrow 0$, the probabilistic measures approach the count-based ones
\item For $\sigma_x \sigma_y \rightarrow \infty$, the probabilistic accuracy approaches   0.5 as all probabilities are approximately 0.5
\item Lag-$l$-measures can easily be obtained by using the $D^l$-difference, i.e., $D^l y_t = y_t - y_{t-l}$
\end{itemize}
